---
title: "STOR 664 Project Report"
output: pdf_document
---

```{r}
library(tidyverse)
library(broom)
library(corrplot)

cap_year <- read_csv("C:/Users/soumy/Downloads/capital_by_position_year.csv")
cap_year <- cap_year %>%
  mutate(
    position = factor(position),
    year_c   = year - min(year)   # e.g., 0 for 2013, 1 for 2014, ...
  )

glimpse(cap_year)
summary(cap_year)
```


```{r}
ggplot(cap_year, aes(x = year, y = cap_pct_lg,
                     color = position, group = position)) +
  geom_line() +
  geom_point(alpha = 0.7) +
  scale_x_continuous(breaks = sort(unique(cap_year$year))) +
  labs(
    title = "League-wide salary cap share by position over time",
    x = "Season",
    y = "Share of league cap"
  ) +
  theme_minimal()
```
The above figure shows how much of the league’s salary cap each position gets from 2013 to 2024. Defensive line (DL) and offensive line (OL) usually get the biggest shares, followed by defensive backs (DB) and wide receivers (WR). Kicker (K), punter (P), and long snapper (LS) always get very small amounts. Most positions do not change very much over time, but there are a few patterns: OL and WR seem to get a little more cap share over the years, while DB and running back (RB) seem to get a little less.
```{r}
ggplot(cap_year, aes(x = year, y = draft_pct_lg,
                     color = position, group = position)) +
  geom_line() +
  geom_point(alpha = 0.7) +
  scale_x_continuous(breaks = sort(unique(cap_year$year))) +
  labs(
    title = "League-wide draft capital share by position over time",
    x = "Season",
    y = "Share of league draft capital"
  ) +
  theme_minimal()
```
The above figure shows a similar plot, but now for how draft picks are used. Compared to cap spending, the draft shares jump around more from year to year. This makes sense because draft choices depend on how strong each draft class is and what teams need that year. Defensive backs (DB), defensive line (DL), running backs (RB), and wide receivers (WR) usually get the most draft capital, while specialists like kicker (K), punter (P), and long snapper (LS) get very little.
```{r}
# Step 4: Full model for cap shares
formula_cap <- cap_pct_lg ~ position * year_c

model_cap <- lm(formula_cap, data = cap_year)
summary(model_cap)
```

\section{Big regression model}

```{r}
# Step 5: Full model for draft shares
formula_draft <- draft_pct_lg ~ position * year_c

model_draft <- lm(formula_draft, data = cap_year)
summary(model_draft)

# Per-position simple time trends for draft share
draft_slopes <- cap_year %>%
  group_by(position) %>%
  do(tidy(lm(draft_pct_lg ~ year_c, data = .))) %>%
  ungroup() %>%
  filter(term == "year_c") %>%
  arrange(desc(estimate)) %>%
  mutate(
    slope_pct_points = 100 * estimate
  )

draft_slopes_report <- draft_slopes %>%
  mutate(
    estimate = round(estimate, 5),
    slope_pct_points = round(slope_pct_points, 3),
    p.value = signif(p.value, 3)
  )

draft_slopes_report
```

\section{Diagnostics}

```{r}
# Step 6: Diagnostics for cap model
par(mfrow = c(2, 2))
plot(model_cap)
par(mfrow = c(1, 1))

# Step 6b: Diagnostics for draft model
par(mfrow = c(2, 2))
plot(model_draft)
par(mfrow = c(1, 1))

```
